---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Intro

- You can make your own functions
- Think of things you do a lot and don't want to have to remember exact syntax

# Automatically viewing snippets of data in Excel

- https://twitter.com/brodriguesco/status/1447468259725434886

## Arguments

```{r}
library(tidyverse)
library(fs)

show_in_excel <- function(data) {
  
  csv_file <- str_glue("{tempfile()}.csv")
  
  write_csv(x = data,
            file = csv_file,
            na = "")
  
  file_show(path = csv_file)
  
}

show_in_excel(penguins)
```

## Can use function in pipe too

```{r}
library(tidyverse)

penguins %>% 
  filter(island == "Torgersen") %>% 
  show_in_excel()
```

# ggplot theme

- Defaults for arguments
- Using if statements
- What gets returned

```{r}
theme_dk <- function(show_grid_lines = TRUE,
                     show_axis_titles = TRUE,
                     theme_font = "Inter") {
  
  theme_dk <- theme_minimal(base_family = theme_font) +
    theme(panel.grid.minor = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 12,
                                   color = "grey50"),
          axis.title.x = element_text(margin = margin(t = 10)),
          axis.title.y = element_text(margin = margin(r = 10)),
          axis.text = element_text(size = 12,
                                   color = "grey50"),
          plot.title.position = "plot",
          plot.title = element_text(size = 20,
                                    face = "bold",
                                    margin = margin(b = 8)),
          plot.subtitle = element_text(size = 14,
                                       color = "grey50"),
          legend.text = element_text(size = 12),
          legend.position = "top")
  
  if (show_grid_lines == FALSE) {
    
    theme_dk <- theme_dk +
      theme(panel.grid.major = element_blank())
    
  }
  
  if (show_axis_titles == FALSE) {
    
    theme_dk <- theme_dk +
      theme(axis.title = element_blank(),
            axis.title.x = element_blank(),
            axis.title.y = element_blank())
    
  }
  
  theme_dk
  
}
```

```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm,
             y = bill_depth_mm,
             color = island)) +
  geom_point() +
  labs(title = "A histogram of bill length and bill depth",
       subtitle = "Data from palmerpenguins package",
       x = "Bill Length",
       y = "Bill Depth",
       color = NULL) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  scale_color_viridis_d() +
  theme_dk()
```


```{r}
penguins %>% 
  count(island) %>% 
  ggplot(aes(x = island,
             y = n)) +
  geom_col() +
  labs(title = "Number of penguins on each island",
       subtitle = "Data from palmerpenguins package") +
  theme_dk(show_grid_lines = FALSE,
           show_axis_titles = FALSE)

```

# Automatically pulling in and formatting race/ethnicity data from ACS

- Using ... for additional arguments

```{r}
library(janitor)

get_acs_race_ethnicity <- function(...) {
  
  get_acs(...,
          variables = c("White" = "B03002_003",
                        "Black/African American" = "B03002_004",
                        "American Indian/Alaska Native" = "B03002_005",
                        "Asian" = "B03002_006",
                        "Native Hawaiian/Pacific Islander" = "B03002_007",
                        "Other race" = "B03002_008",
                        "Multi-Race" = "B03002_009",
                        "Hispanic/Latino" = "B03002_012")) %>%
    clean_names() %>%
    set_names(c("geoid", "geography", "population_group", "estimate", "moe"))
  
}

get_acs_race_ethnicity(geography = "county",
                       year = 2020)
```


# Conclusion

- It's tempting to see others' functions as real and your as not, but they're all the same thing
- Next chapter will show how to create your own personal package to help you keep functions at the ready
