---
title: "Penguins Report"
format:
  revealjs: 
    theme: theme.scss
execute: 
  echo: false
  warning: false
  message: false
---

# Introduction

```{r}
library(tidyverse)
```

```{r}
penguins <- read_csv("https://raw.githubusercontent.com/rfortherestofus/r-without-statistics/main/data/penguins-2008.csv")
```

We are writing a report about the **Palmer Penguins**. These penguins are *really* amazing. There are three species:

- Adelie
- Gentoo
- Chinstrap

```{css}
.center-slide {
	text-align: center;
}
```

## Bill Length {.center-slide}

We can make a histogram to see the distribution of bill lengths.

```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram() +
  theme_minimal()
```

```{r}
average_bill_length <- penguins %>% 
  summarize(avg_bill_length = mean(bill_length_mm,
                                   na.rm = TRUE)) %>% 
  pull(avg_bill_length)
```

The chart shows the distribution of bill lengths. The average bill length is `r average_bill_length` millimeters.

---

:::: {.columns}

::: {.column width="50%"}
```{r}
penguins %>% 
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram() +
  theme_minimal()
```

:::

::: {.column width="50%"}
```{r}
penguins %>% 
  ggplot(aes(x = bill_depth_mm)) +
  geom_histogram() +
  theme_minimal()
```

:::

::::

## Penguins {background-image="penguins.jpg"}