---
title: "Interview with Tom Mock"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(gt)
tuesdata <- tidytuesdayR::tt_load(2020, "36")
```

# Your Background

Customer enablement lead at RStudio

## When did you start using R?

## Why did you switch to R? 

## What changed when you moved to R? 

## You've said that the most difficult thing when you were learning R was reading in data. Why was that? 

## You now run Tidy Tuesday. In what ways would your learning journey have been different if Tidy Tuesday had existed when you were learning R? 

# Tables

## Why do you care so much about tables? 

## What makes R well suited to making tables?

## Talk about the table package landscape. Which packages do you find most useful? Why?

# Code walkthrough

```{r}
country_sel <- c("China", "India", "United States", "Indonesia", "Mexico", "Pakistan")

yield_data <- tuesdata$key_crop_yields %>% 
  janitor::clean_names() %>% 
  rename_with(~str_remove(., "_tonnes_per_hectare")) %>% 
  select(entity:beans, -code) %>% 
  pivot_longer(cols = wheat:beans, names_to = "crop", values_to = "yield") %>% 
  rename(Country = entity)
```

## Offset heads from body

```{r}
potato_data <- yield_data %>% 
  filter(Country %in% country_sel, crop == "potatoes", year %in% c(2013:2016)) %>% 
  filter(crop == "potatoes") %>% 
  pivot_wider(names_from = year, values_from = "yield")
```

### Poor example

```{r}
potato_tb <- potato_data %>% 
  gt() %>% 
  cols_hide(vars(crop)) %>% 
  opt_table_lines(extent = "none") %>% 
  fmt_number(
    columns = 3:6,
    decimals = 2
  )

potato_tb
```

```{r}

```


### Improved example

```{r}
potato_tb %>% 
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  ) %>% 
  opt_table_lines(extent = "default") %>%
  tab_options(
    column_labels.border.top.color = "white",
    column_labels.border.top.width = px(3),
    column_labels.border.bottom.color = "black",
    table_body.hlines.color = "white",
    table.border.bottom.color = "white",
    table.border.bottom.width = px(3)
  ) %>% 
  tab_source_note(md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br>**Inspiration**: @jschwabish"))
```


## Use subtle dividers

```{r}
rule2_data <- yield_data %>% 
  filter(Country %in% country_sel, crop == "potatoes", year %in% c(2007:2016)) %>% 
  filter(crop == "potatoes") %>% 
  select(-crop) %>% 
  pivot_wider(names_from = year, values_from = "yield") %>% 
  rowwise() %>% 
  mutate(
    avg_07_11 = mean(`2007`:`2011`),
    .before = `2012`
    ) %>% 
  mutate(
    avg_12_16 = mean(`2012`:`2016`)
  ) %>% 
  ungroup()
```

### Poor example

```{r}
rule2_data %>% 
  gt(
    rowname_col = "Country"
  ) %>% 
  cols_label(
    avg_07_11 = "Avg.",
    avg_12_16 = "Avg."
  ) %>% 
  cols_width(
    1 ~ px(125)
  ) %>% 
  fmt_number(
    columns = 2:last_col()
  ) %>% 
  tab_style(
    style = cell_borders(
      side = "all",
      color = "grey",
      weight = px(1),
      style = "solid"
    ),
    locations = list(
      cells_body(
        everything()
        ),
      cells_column_labels(
        everything()
      )
      )
  ) %>% 
  grand_summary_rows(
    columns = 2:last_col(),
    fns = list(
      "Average" = ~mean(.)
    ),
    formatter = fmt_number
  )


```

### Improved example

```{r}
rule2_data %>% 
  add_row(
    rule2_data %>% 
      summarize(
        across(where(is.double), 
               list(Average = mean),
               .names = "{col}")
      ) %>% 
      mutate(Country = "Average")
  ) %>% 
  gt() %>% 
  cols_label(
    avg_07_11 = "Avg.",
    avg_12_16 = "Avg."
  ) %>%
  fmt_number(
    columns = 2:last_col()
  ) %>% 
  tab_style(
    style = cell_fill(
      color = "lightgrey"
    ),
    locations = list(
      cells_body(
        columns = vars(avg_07_11, avg_12_16)
        ),
      cells_column_labels(
        columns = vars(avg_07_11, avg_12_16)
      )
      )
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "black",
      weight = px(2)
    ),
    locations = cells_body(
      columns = everything(),
      rows = Country == "Average"
    )
  ) %>% 
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  ) %>% 
  tab_options(
    column_labels.border.top.color = "black",
    column_labels.border.top.width = px(3),
    column_labels.border.bottom.color = "black"
  )
```



## Alignment

### Right align numbers and heads

```{r}
# Prep data
rule3_data <- yield_data %>% 
  filter(Country == "United States", year %in% c(2016)) %>% 
  mutate(crop = str_to_title(crop)) %>% 
  pivot_wider(names_from = year, values_from = "yield") %>% 
  arrange(crop) %>% 
  select(-Country, Crop = crop)
```

```{r}
rule3_data %>% 
  mutate(`Center align` = `2016`,
             `Right align` = `2016`) %>%
  rename(`Left align` = 2) %>% 
  gt() %>% 
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  ) %>% 
  fmt_number(
    columns = 2:4
  ) %>% 
  cols_align(align = "left",
             columns = 2) %>% 
  cols_align(align = "center",
             columns = 3) %>% 
  cols_align(align = "right",
             columns = 4) %>% 
  tab_options(
    column_labels.border.top.color = "white",
    column_labels.border.top.width = px(3),
    column_labels.border.bottom.color = "black",
    table_body.hlines.color = "white",
    table.border.bottom.color = "white",
    table.border.bottom.width = px(3)
  ) %>% 
  opt_table_font(font = "Courier New")
```


### Left align text and heads

```{r}
country_names <- c(
  "British Virgin Islands",
  "Cayman Islands",
  "Democratic Republic of Congo",
  "Luxembourg", 
  "United States",
  "Germany",
  "New Zealand",
  "Costa Rica",
  "Peru"
)

tibble(
  right = country_names,
  center = country_names,
  left = country_names
) %>% 
  gt()  %>% 
  cols_align(align = "left",
             columns = 3) %>% 
  cols_align(align = "center",
             columns = 2) %>% 
  cols_align(align = "right",
             columns = 1) %>% 
  cols_width(
    everything() ~ px(250)
  ) %>% 
  tab_options(
    column_labels.border.top.color = "white",
    column_labels.border.top.width = px(3),
    column_labels.border.bottom.color = "black",
    column_labels.font.weight = "bold",
    table_body.hlines.color = "white",
    table.border.bottom.color = "white",
    table.border.bottom.width = px(3),
    data_row.padding = px(3)
  ) %>% 
  cols_label(
    right = md("Right aligned and<br>hard to read"),
    center = md("Centered and<br>even harder to read"),
    left = md("Left-aligned and<br>easiest to read")
  ) %>% 
  tab_source_note(md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br>**Inspiration**: @jschwabish"))
```


## Select appropriate level of precision

```{r}
yield_data %>% 
  filter(Country %in% country_sel, crop == "potatoes", year %in% c(2016)) %>% 
  select(Country, yield) %>% 
  mutate(few = yield, right = yield) %>% 
  gt() %>% 
  fmt_number(
    columns = vars(few),
    decimals = 0
  ) %>%
  fmt_number(
    columns = vars(right),
    decimals = 1
  ) %>%
  cols_label(
    yield = md("Too many<br>decimals"),
    few = md("Too few<br>decimals"),
    right = md("About<br>right")
  ) %>% 
  tab_source_note(md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br>**Inspiration**: @jschwabish"))
```


## Highlight outliers

```{r}
rule8_data <- yield_data %>% 
  filter(Country %in% country_sel, crop == "potatoes", year %in% 2009:2017) %>% 
  group_by(Country) %>% 
  mutate(pct_change = (yield/lag(yield)-1)*100) %>% 
  ungroup() %>% 
  filter(between(year, 2010, 2016)) %>% 
  select(Country, year, pct_change) %>% 
  pivot_wider(names_from = year, values_from = pct_change)
```

```{r}
rule8_tb <- rule8_data %>% 
  gt() %>% 
  fmt_number(2:last_col()) %>% 
  cols_label(
    Country = ""
  ) %>% 
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  ) %>% 
  tab_options(
    column_labels.border.top.color = "white",
    column_labels.border.top.width = px(3),
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(3),
    table_body.hlines.color = "white",
    table.border.bottom.color = "black",
    table.border.bottom.width = px(3)
  ) %>% 
  cols_width(vars(Country) ~ px(125),
             2:last_col() ~ px(75))
```

### Without highlighting

```{r}
rule8_tb %>% 
  tab_source_note(md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br>**Inspiration**: @jschwabish"))
```

### With highlighting

```{r}
rule8_tb %>% 
  tab_style(
    style = list(
      cell_fill(color = "red"),
      cell_text(color = "white")
    ),
    locations = list(
      cells_body(
        columns = 2,
        rows = `2010` < 0
    ),
    cells_body(
      columns = 3,
      rows = `2011` < 0
    ),
    cells_body(
      columns = 4,
      rows = `2012` < 0
    ),
    cells_body(
      columns = 5,
      rows = `2013` < 0
    ),
    cells_body(
      columns = 6,
      rows = `2014` < 0
    ),
    cells_body(
      columns = 7,
      rows = `2015` < 0
    ),
    cells_body(
      columns = 8,
      rows = `2016` < 0
    )
  )
  ) %>% 
  tab_source_note(md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br>**Inspiration**: @jschwabish"))
```



## Group similar data


```{r}
rule9_data <- yield_data %>% 
  filter(Country %in% country_sel[-5], year %in% c(2015, 2016),
         crop %in% c("wheat", "potatoes", "rice", "soybeans"),
         !is.na(yield)) %>% 
  pivot_wider(names_from = year, values_from = yield) %>% 
  rowwise() %>% 
  mutate(crop = str_to_title(crop),
         pct_change = (`2016`/`2015`-1)*100) %>%
  group_by(Country) %>% 
  arrange(desc(`2015`)) %>% 
  ungroup() 
```

### Bad example


```{r}
rule9_data %>% 
  gt() %>% 
  fmt_number(
    columns = vars(`2015`, `2016`, pct_change)
  ) %>% 
  tab_spanner(columns = vars(`2015`, `2016`),
              label = md("**Yield in Tonnes/Hectare**")) %>%  
  cols_width(
    vars(crop) ~ px(125),
    vars(`2015`, `2016`, pct_change) ~ 100
  ) %>% 
  tab_source_note(md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br>**Inspiration**: @jschwabish"))
```

### gt native grouping

```{r}
rule9_data %>% 
  gt(groupname_col = "Country") %>% 
  tab_stubhead("label") %>% 
  tab_options(
    table.width = px(300)
  ) %>% 
  cols_label(
    crop = "",
    pct_change = md("Percent<br>Change")
  ) %>% 
  fmt_number(
    columns = vars(`2015`, `2016`, pct_change)
  ) %>% 
  tab_style(
    style = cell_text(color = "black", weight = "bold"),
    locations = list(
      cells_row_groups(),
      cells_column_labels(everything())
    )
  ) %>% 
  tab_spanner(columns = vars(`2015`, `2016`),
              label = md("**Yield in Tonnes/Hectare**")) %>%  
  cols_width(
    vars(crop) ~ px(125),
    vars(`2015`, `2016`, pct_change) ~ 100
  ) %>% 
  tab_options(
    row_group.border.top.width = px(3),
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "black",
    table_body.hlines.color = "white",
    table.border.top.color = "white",
    table.border.top.width = px(3),
    table.border.bottom.color = "white",
    table.border.bottom.width = px(3),
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(2)
  ) %>% 
  tab_source_note(md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br>**Inspiration**: @jschwabish"))
```
### Remove duplicate data

```{r}
rule9_data %>% 
  arrange(Country) %>% 
  gt() %>% 
  cols_label(
    Country = "",
    crop = "Crop",
    pct_change = md("Percent<br>Change")
  ) %>% 
  tab_spanner(columns = vars(`2015`, `2016`),
              label = md("**Yield in Tonnes/Hectare**")) %>% 
  fmt_number(
    columns = vars(`2015`, `2016`, pct_change)
  ) %>% 
  text_transform(
    locations = cells_body(
      columns = vars(Country),
      rows = crop != "Potatoes"
    ),
    fn = function(x){
      paste0("")
    }
  ) %>% 
  tab_style(
    style = cell_text(color = "black", weight = "bold"),
    locations = list(
      cells_row_groups(),
      cells_column_labels(everything())
    )
  ) %>% 
  cols_width(
    vars(Country, crop) ~ px(125),
    vars(`2015`, `2016`, pct_change) ~ 100
  ) %>% 
  tab_options(
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(2),
    table_body.hlines.color = "white",
    table.border.top.color = "white",
    table.border.top.width = px(3),
    table.border.bottom.color = "white",
    table.border.bottom.width = px(3),
  ) %>% 
  tab_source_note(md("**Table**: @thomas_mock | **Data**: OurWorldInData.org<br>**Inspiration**: @jschwabish"))
```



# Wrap Up

## You've developed `gtextras`. What needs does this package fill?