---
title: "Interview with Kyle Walker"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidycensus)
library(tidyverse)
library(janitor)
```

# Background

## Tell me a bit about your background. 

## What were you using before you used R? 

## When and why did you start using R?

## What changed when you moved to R? 

# Background on tidycensus

## Where did the idea for tidycensus come from. 

## How did you work with Census data before you created tidycensus?

## What need does tidycensus meet? 

# Code walkthrough

## See all variables

```{r}
census_2020_vars <- load_variables(2020, "pl")
```

## Work with Census data (Hispanic/Latino population by state)

```{r}
get_decennial(geography = "state", 
              variables = "P2_002N",
              year = 2020)
```

### Name your variables

```{r}
get_decennial(geography = "state", 
              variables = c(hispanic_latino_population = "P2_002N"),
              year = 2020)
```

### Calculate percentages 

```{r}
get_decennial(geography = "state", 
              variables = c(hispanic_latino_population = "P2_002N",
                            not_hispanic_latino_population = "P2_003N"),
              year = 2020) %>% 
  group_by(NAME) %>% 
  mutate(pct = value / sum(value))  %>% 
  ungroup() %>% 
  filter(variable == "hispanic_latino_population") %>% 
  arrange(desc(pct))
```

### Calculate percentages with a summary variable

```{r}
get_decennial(geography = "state", 
              variables = c(hispanic_latino_population = "P2_002N"),
              summary_var = "P2_001N",
              year = 2020) %>% 
  mutate(pct = value / summary_value) %>% 
  arrange(desc(pct))
```

## ACS data (median age by state)

```{r}
get_acs(
  geography = "state",
  variables = "B01002_001",
  year = 2020
)
```

### Pipe directly into ggplot

```{r}
get_acs(
  geography = "state",
  variables = "B01002_001",
  year = 2020
) %>% 
  ggplot(aes(x = estimate,
             y = NAME)) +
  geom_col()
```

## Get geospatial data

```{r}
get_acs(
  geography = "state",
  variables = "B01002_001",
  year = 2020,
  geometry = TRUE
) 
```

### Map your results

```{r}
get_acs(
  geography = "state",
  variables = "B01002_001",
  year = 2020,
  geometry = TRUE
) %>% 
  filter(NAME != "Alaska" & NAME != "Hawaii" & NAME != "Puerto Rico") %>% 
  ggplot(aes(fill = estimate)) +
  geom_sf() +
  scale_fill_viridis_c()
```


